---
title: "Point Cloud"
author: "Dan Miller"
date: "`r Sys.Date()`"
format: 
  html:
    code-fold: true
editor: visual
---

Tasks:

1.  Assemble point clouds. This involves several subtasks:\

    a)  Create a LAScatalog for each data set. This provides a polygon shapefile showing all las tile locations.\
    b)  Create a lidR las object for the area of interest.\
    c)  Classify ground and water returns (or rely on existing classifications). Potentially filter outliers. Need to come up with an algorithm for that.\
    d)  Las objects include X, Y, and Z data fields. After the points have been classified and filtered, export ground and water returns to a data.table object. Washington state data are in state plane U.S. survey feet. Convert everything to meters. Note that US survey feet are 0.3048006096012192 meters in length.

2.  Generate DTMs from the point clouds (or use the one downloaded from the portal?). Use tin interpolation. Horizontal cell resolution of 1 meter (?).

3.  Code to enforce locally low ground points. See DEM Alignment. However, use a slightly different algorithm. Rank ground returns by elevation below the interpolated DTM surface and create an array with elevation difference and DTM column and row. Starting with the point having the largest-magnitude elevation difference, find the nearest DTM grid point and set its elevation to the ground-return elevation. March through all the points.

4.  Code to align DTMs as described in DEM Alignment.

5.  Try Iterative-Closest-Point algorithm as well, generate DTMs from the resulting shifted point clouds, and compare to results from step 4.

6.  Find landslides

```{r}
#| echo: false
library(data.table)
library(lidR)
library(raster)
library(sf)
library(terra)
library(ggplot2)
library(qgisprocess)
```

Read in the 2006 Lewis acquisition as a las catalog and export the tile locations to a polygon shapefile.

```{r}
path <- "c:/work/data/postmortem/"

datafile <- paste0(path, "lewis_2006/tilepoly.shp")
if (!file.exists(datafile)) {
  lewis_2006 <- readLAScatalog(paste0(path, "lewis_2006/laz")) 
  st_write(st_as_sf(lewis_2006), datafile, append = FALSE)
}

datafile <- paste0(path, "swwa_foothills_2017/tilepoly.shp")
if (!file.exists(datafile)) {
  swwa_2017 <- readLAScatalog(paste0(path, "swwa_foothills_2017/laz"))
  st_write(st_as_sf(swwa_2017), datafile, append = FALSE)
}

datafile <- paste0(path, "southwest_wa_opsw_2019/tilepoly.shp")
if (!file.exists(datafile)) {
  opsw_2019 <- readLAScatalog(paste0(path, "southwest_wa_opsw_2019/laz"))
  st_write(st_as_sf(opsw_2019), datafile, append = FALSE)
}
```

Clip out a region of interest:

```{r}
datafile <- paste0(path, "cluster_1SP.shp")
box <- st_read(datafile)

all_2006 <- lidR::clip_roi(lewis_2006, box) 
lidR::filter_duplicates(all_2006)
ground_2006 <- filter_poi(all_2006, Classification == 2L | Classification == 9L) # keep ground returns and water

all_2017 <- lidR::clip_roi(swwa_2017, box) 
lidR::filter_duplicates(all_2017)
ground_2017 <- filter_poi(all_2017, Classification == 2L | Classification == 9L)

all_2019 <- lidR::clip_roi(opsw_2019, box) 
lidR::filter_duplicates(all_2019)
ground_2019 <- filter_poi(all_2019, Classification == 2L | Classification == 9L)
```

The laz files from the lidar portal are in StatePlane US feet. I haven't figured out how to convert a spatraster from feet to meters in the horizontal units, so that will need to happen with subsequent processing.

```{r}
# This produces a DTM using tin interpolation with 1-meter cell size
dtm_2006 <- lidR::rasterize_terrain(ground_2006, res=3.28083333333333335958, algorithm = tin(), pkg="terra")

# This gives point density in points per square foot
#den_2006 <- lidR::rasterize_density(ground_2006, res=3.28083333333333335958)
#den_2006 <- den_2006 * 10.76386736 # convert to points per cell

writeRaster(dtm_2006, paste0(path, "dtm_2006.tif"), overwrite=TRUE)
#writeRaster(den_2006, paste0(path, "den_2006.tif"), overwrite=TRUE)

dtm_2017 <- lidR::rasterize_terrain(ground_2017, res=3.28083333333333335958, algorithm = tin(), pkg="terra")

# This gives point density in points per square foot
#den_2017 <- lidR::rasterize_density(ground_2017, res=3.28083333333333335958)
#den_2017 <- den_2017 * 10.76386736 # convert to points per cell

writeRaster(dtm_2017, paste0(path, "dtm_2017.tif"), overwrite=TRUE)
#writeRaster(den_2017, paste0(path, "den_2017.tif"), overwrite=TRUE)

#dtm_2019 <- lidR::rasterize_terrain(ground_2019, res=3.28083333333333335958, algorithm = tin(), pkg="terra")

# This gives point density in points per square foot
#den_2019 <- lidR::rasterize_density(ground_2019, res=3.28083333333333335958)
#den_2019 <- den_2019 * 10.76386736 # convert to points per cell

#writeRaster(dtm_2019, paste0(path, "dtm_2019.tif"), overwrite=TRUE)
#writeRaster(den_2019, paste0(path, "den_2019.tif"), overwrite=TRUE)
```

This will export the x, y, z values to a csv file and convert everything to meters (I hope)

```{r}
xyz_2006 <- as.data.table(list(ground_2006$X, ground_2006$Y, ground_2006$Z))
xyz_2006 <- xyz_2006[, .(X=V1*0.3048006096012192, Y=V2*0.3048006096012192, Z = V3*0.3048006096012192)]
write.csv(xyz_2006,paste0(path, "xyz_2006.csv"))

xyz_2017 <- as.data.table(list(ground_2017$X, ground_2017$Y, ground_2017$Z))
xyz_2017 <- xyz_2017[, .(X=V1*0.3048006096012192, Y=V2*0.3048006096012192, Z = V3*0.3048006096012192)]
write.csv(xyz_2017,paste0(path, "xyz_2017.csv"))

xyz_2019 <- as.data.table(list(ground_2019$X, ground_2019$Y, ground_2019$Z))
xyz_2019 <- xyz_2019[, .(X=V1*0.3048006096012192, Y=V2*0.3048006096012192, Z = V3*0.3048006096012192)]
write.csv(xyz_2019,paste0(path, "xyz_2019.csv"))
```

The align.exe fortran code appears to work as intended. An interesting result: the DTMs constructed using lidR from the point clouds in the code chunks above align more closely than the DTMs downloaded from the lidar portal. The plan, maybe:

1.  Identify overlapping laz tiles from the 2006, 2017, and 2019 lidar that are within post-mortem cluster polygons, build las catalogs, and clip out an area of interest, and build DTMs for each lidar acquisition. Have the DTMs align exactly by x, y corners and cellsize.

2.  Modify align.exe to divide the DTM into a set of overlapping tiles of specified size. Find dx, dy, dz for each tile. Assign those shifts to the tile midpoint. Once shifts are found for all tiles, interpolate to get a dx, dy, dz for each DTM grid point and each ground return. Use the shifted ground returns to build a new DTM.

3.  Create a difference raster.
